---
title: "Exploring eBird data in Canada in 2025"
format: html
---

```{r}
#| include: false
library(auk)
library(ggplot2)
library(dplyr)
library(tidyr)
library(lubridate)

theme_set(theme_bw())
```

Loading the data. We compute some time related variables and remove observations from ebird specialized protocols.


```{r}
dat_s <- readRDS("data/dat_s.rds") |>
  # Calculate fields
  mutate(year = year(observation_date), 
         month = month(observation_date, label = F), 
         month_lab = month(observation_date, label = T),
         week = week(observation_date), 
         weekday = wday(observation_date, label = T),
         time_observations_started_hms = hms(time_observations_started),
         hour = hour(time_observations_started_hms),
         minutes = minutes(time_observations_started_hms)
        ) |>
  # Right off the bat, remove observation from specialized protocols
  filter(!(observation_type %in% c(
    "eBird Pelagic Protocol", "Nocturnal Flight Call Count", "Banding", "Area"
  )))

dat_s |> str()
```

Simple checks.

```{r}
nrow(dat_s) # 1 246 952
dat_s$all_species_reported |> all() # TRUE, only complete checklists
dat_s$checklist_id |> unique() |> length() # 1 246 952 => same
```

### Numbers of observers

```{r}

dat_s$number_observers |> is.na() |> sum() # 3549 missing number of observers
dat_s$number_observers |> mean(na.rm = T) # 1.458653
dat_s$number_observers |> median(na.rm = T) # 1
dat_s$number_observers |> table()
dat_s$number_observers |> sort(decreasing = TRUE) |> head()
# dat_s |> filter(number_observers >= 100) |> View()
```

How do number of observers change with time? Some checklists have insanely high number of observers, probably due to input errors. We filter to <= 100 obserbers.

```{r}
dat_s_obs_m <- dat_s |>
  filter(number_observers<=100) |>
  select(checklist_id, month, number_observers) |>
  group_by(month) |>
  summarize(mean_number_obs = mean(number_observers, na.rm = T), 
            n_samples = n(),
            sd_number_obs = sd(number_observers, na.rm = T)) |>
  mutate(se = sd_number_obs/sqrt(n_samples),
         mean_number_obs_upper = mean_number_obs+1.96*se,
         mean_number_obs_lower = mean_number_obs-1.96*se)

dat_s_obs_w <- dat_s |>
  filter(number_observers<=100) |>
  select(checklist_id, week, number_observers) |>
  group_by(week) |>
  summarize(mean_number_obs = mean(number_observers, na.rm = T), 
            n_samples = n(),
            sd_number_obs = sd(number_observers, na.rm = T)) |>
  mutate(se = sd_number_obs/sqrt(n_samples),
         mean_number_obs_upper = mean_number_obs+1.96*se,
         mean_number_obs_lower = mean_number_obs-1.96*se)
```

Now we plot:

```{r}
dat_s_obs_m |>
  ggplot(aes(x = month, y = mean_number_obs, 
             ymin = mean_number_obs_lower, 
             ymax = mean_number_obs_upper)) +
  geom_point() + 
  geom_errorbar()
```

```{r}
dat_s_obs_w |>
  ggplot(aes(x = week, y = mean_number_obs, 
             ymin = mean_number_obs_lower, 
             ymax = mean_number_obs_upper)) +
  geom_point() + 
  geom_errorbar()
```

### Observation date

Mean and median.

```{r}
dat_s$observation_date |> mean() # "2025-06-22"
dat_s$observation_date |> median() # "2025-06-10"
```

Simple viz.

```{r}
ggplot(dat_s) +
  geom_histogram(aes(month_lab), stat="count")
ggplot(dat_s) +
  geom_histogram(aes(week), stat="count")
ggplot(dat_s) +
  geom_histogram(aes(weekday), stat="count")
ggplot(dat_s) +
  geom_histogram(aes(weekday), stat="count")
```

Then we group.

```{r}
dat_s_date_m <- dat_s |> 
  group_by(month, weekday) |> 
  summarise(weekday_n = n())
dat_s_date_wk <- dat_s  |> 
  group_by(week, weekday) |> 
  summarise(weekday_n = n())
```

More viz.

```{r}
ggplot(dat_s_date_m, aes(fill=weekday, y=weekday_n, x=month)) + 
    geom_bar(position="stack", stat="identity")
ggplot(dat_s_date_wk, aes(fill=weekday, y=weekday_n, x=week)) + 
    geom_bar(position="stack", stat="identity")

ggplot(dat_s_date_m, aes(color=weekday, y=weekday_n, x=month)) + 
    geom_point() +
    geom_smooth(se = F, method = "gam", formula = y ~ s(x, k=12))
ggplot(dat_s_date_wk, aes(color=weekday, y=weekday_n, x=week)) +
    geom_point() +
    geom_smooth(se = F, method = "gam", formula = y ~ s(x, k=12))
```

### NAs exploration

We explore distributions of NA values in the dataset.

```{r}
is_not_na <- \(x) !is.na(x)

dat_s$effort_area_ha |> is_not_na() |> sum() # 1159
dat_s$effort_distance_km |> is.na() |> sum() # 442375

dat_s$observation_type |> is.na() |> sum() # None
dat_s$observation_type |> table(useNA = "ifany")
dat_s |>
  mutate(effort_distance_is_na = is.na(effort_distance_km)) |>
  select(effort_distance_is_na, observation_type) |> table(useNA = "ifany")

dat_s$protocol_name |> is.na() |> sum()
dat_s$protocol_name |> table(useNA = "ifany")
dat_s |>
  mutate(effort_distance_is_na = is.na(effort_distance_km)) |>
  select(effort_distance_is_na, protocol_name) |> table(useNA = "ifany")

dat_s$protocol_code |> is.na() |> sum() 
dat_s$protocol_code |> table(useNA = "ifany")
dat_s |>
  mutate(effort_distance_is_na = is.na(effort_distance_km)) |>
  select(effort_distance_is_na, protocol_code) |> table(useNA = "ifany")

dat_s$project_names |> is.na() |> sum()
dat_s$project_names |> table(useNA = "ifany")
dat_s |>
  mutate(effort_distance_is_na = is.na(effort_distance_km)) |>
  select(effort_distance_is_na, project_names) |> table(useNA = "ifany")

dat_s$project_identifiers |> is.na() |> sum()
dat_s$project_identifiers |> table(useNA = "ifany")
dat_s |>
  mutate(effort_distance_is_na = is.na(effort_distance_km)) |>
  select(effort_distance_is_na, project_identifiers) |> table(useNA = "ifany")
```

### Checklist duration

```{r}
dat_s |> 
  ggplot(aes(x = duration_minutes)) +
  geom_histogram(binwidth = 5) +
  facet_wrap(~month_lab)
```

Some grouping.

```{r}
dat_s_dur_m <- dat_s |>
  select(checklist_id, month, duration_minutes) |>
  group_by(month) |>
  summarize(mean_dur = mean(duration_minutes, na.rm = T), 
            n_samples = n(),
            sd_dur = sd(duration_minutes, na.rm = T)) |>
  mutate(se = sd_dur/sqrt(n_samples),
         mean_dur_upper = mean_dur+1.96*se,
         mean_dur_lower = mean_dur-1.96*se)

dat_s_dur_m_w <- dat_s |>
  select(checklist_id, week, duration_minutes) |>
  group_by(week) |>
  summarize(mean_dur = mean(duration_minutes, na.rm = T), 
            n_samples = n(),
            sd_dur = sd(duration_minutes, na.rm = T)) |>
  mutate(se = sd_dur/sqrt(n_samples),
         mean_dur_upper = mean_dur+1.96*se,
         mean_dur_lower = mean_dur-1.96*se)
```

Basic viz.

```{r}
dat_s_dur_m |>
  ggplot(aes(x = month, y = mean_dur, 
             ymin = mean_dur_upper, 
             ymax = mean_dur_lower)) +
  geom_point() + 
  geom_errorbar()

dat_s_dur_m_w |>
  ggplot(aes(x = week, y = mean_dur, 
             ymin = mean_dur_upper, 
             ymax = mean_dur_lower)) +
  geom_point() + 
  geom_errorbar()
```

### Time observation started

```{r}
dat_s$time_observations_started |> is.na() |> sum()
dat_s |>
  ggplot(aes(x = hour)) +
  geom_histogram(binwidth = 1)
```

### Looking into effort

```{r}
# plot(dat_s$hour, dat_s$duration_minutes)
cor.test(dat_s$hour, dat_s$duration_minutes)
```

Effort calculation. In the [ebird guide for best practices](https://ebird.github.io/ebird-best-practices/ebird.html#sec-ebird-effort), it is said:

> "Based on our experience working with these data in the context of eBird Status and Trends,  we suggest restricting checklists to traveling or stationary counts less than 6 hours in duration and 10 km in length, at speeds below 100km/h, and with 10 or fewer observers [...]"

In comparison, ebird status and trends filter according to looser criteria:

> Submitted as of 10 February 2024 //
> Observation dates from 1 January 2009 through 31 December 2023 //
> Complete checklists (all bird species detected and identified were included) //
> The primary checklist in a shared checklist //
> Checklists that used the generic traveling or stationary protocols (i.e., not incidental protocol) //
> If traveling checklists, were not longer than 10 kilometers (up to 30 km for species modeled only over ocean) //
> Not longer in duration than 24 hours //
> Contained information on: start time, duration, protocol, number of observers, and distance traveled. //
> Counts of species were available (i.e., a species was not just noted as ‘present’)

Let's follow and highlight the subsetting according to best practcies first, to see how much data we are removing.


```{r}
dat_s_effort <- dat_s |>
  # NAs
  filter(!is.na(duration_minutes)) |>
  filter(!is.na(effort_distance_km)) |>
  filter(!is.na(number_observers)) |>

  # Calculate effort "speed"
  mutate(effort_speed_kmh = effort_distance_km/(duration_minutes/60)) |>
  # ifelse(observation_type == "Traveling", 
  # (effort_distance_km/(duration_minutes/60)), NA)) |>

  # Categorize effort_speed
  mutate(effort_speed_kmh_cat = case_when(
    effort_speed_kmh == 0 ~ "Stationary",
    effort_speed_kmh < 4 ~ "(-) Avg Walking",
    effort_speed_kmh <= 6 ~ "Avg Walking",
    effort_speed_kmh > 6 ~ "(+) Avg Walking"
  )) |>

  # Highlight
  mutate(to_keep = (
    (observation_type %in% c("Stationary", "Traveling")) &
      (number_observers <= 10) &
      (duration_minutes <= (60*6)) &
      (effort_distance_km <= 10) &
      (effort_speed_kmh <= 100)
  ))
  
  # mutate(weight = dnorm(effort_speed, mean = 5, sd = 1.5)) |>
  # slice_sample(n = 500000, weight_by = weight, replace = FALSE)
  # filter((effort_speed > 1) & (effort_speed < 12))
  # mutate(duration_minutes_sc = scale(duration_minutes),
  #        effort_distance_km_sc = scale(effort_distance_km),
  #        number_observers_sc = scale(number_observers)) |>

max(dat_s_effort$effort_speed_kmh, na.rm = T)
# dat_s_effort |> 
#   filter(effort_speed == 20) |> View()
```

```{r}
table(dat_s_effort$observation_type, dat_s_effort$effort_speed_kmh == 0)
```

```{r}
dat_s_effort |>
  # filter(to_keep) |>
ggplot() +
  geom_point(aes(
    x = effort_distance_km, 
    y = duration_minutes/60,
    alpha = to_keep, 
    color = to_keep)) +
  scale_color_manual(values = c(
    "TRUE" = "#42b386",
    "FALSE" = "#d69c5f")
  ) +
  scale_alpha_manual(values = c(
    "TRUE" = 0.8,
    "FALSE" = 0.1)) +
  # scale_color_viridis_c() +
  labs(x = "Effort Distance (km)", y = "Duration (hr)")
```

```{r}
dat_s_effort |> filter(observation_type == "Stationary") |> 
  select(observation_type, effort_speed_kmh_cat) |> table()
dat_s_effort |> filter(observation_type == "Stationary") |>
  select(checklist_id, observation_type, effort_speed_kmh)
```

```{r}
dat_s_effort |>
  filter(to_keep) |>
  arrange(desc(observation_type)) |>
ggplot() +
  geom_point(aes(
    x = effort_distance_km, 
    y = (duration_minutes/60),
    fill = effort_speed_kmh_cat,
    shape = observation_type,
    color = observation_type,
  )) +
  scale_fill_manual(values = c(
    "Stationary"      = "#e0b1cb40",
    "(-) Avg Walking" = "#be95c440",
    "Avg Walking"     = "#5e548e40",
    "(+) Avg Walking" = "#9f86c040"
  )) +
  scale_shape_manual(values = c(
    "Traveling"  = 21,
    "Stationary" = 23
  ))+
  scale_color_manual(values = c(
    "Traveling"  = "#ffffff00",
    "Stationary" = "#c81d25"
  )) +
  # scale_alpha_manual(values = c(
  #   "TRUE" = 0.8,
  #   "FALSE" = 0.1)) +
  labs(x = "Effort Distance (km)", y = "Duration (hr)", color = "Obs Type", shape = "Obs Type", fill = "Speed Cat")
```

```{r}
plot(density(dat_s_effort$effort_speed_kmh))

cor.test(dat_s_effort$effort_distance_km, dat_s_effort$duration_minutes)

dat_s_effort_calc <- dat_s_effort |>
  mutate(effort = (effort_distance_km * duration_minutes) * number_observers)

dat_s_effort_calc$effort |> max()
dat_s_effort_calc$effort |> min()
dat_s_effort_calc$effort |> mean()
dat_s_effort_calc$effort |> median()

dat_s_effort_calc |>
  ggplot(aes(x = log(effort))) +
  geom_histogram(bins = 100)

dat_s_effort_calc |>
  filter(to_keep) |>
  ggplot(aes(x = log(effort))) +
  geom_histogram(bins = 100) +
  facet_wrap(~effort_speed_kmh_cat, scales = "free")
```